Structured Versus Decorated Cospans

### 2020-12-02 cospans.tex
finished with all structure for theorem 2.2

@@

(B) edge node [above] {$j_m$} (F)
(D) edge node [above] {$j_{n}$} (F)
(F) edge node [left] {$\psi$} (G)

@@

is formed as follows: for three composable horizontal 1-cells $M_1=(a\to m_1\leftarrow b,x_1\in F(m_1))$, $M_2=(b\to m_2\leftarrow c,x_2\in F(m_2)$ and $M_3=(c\to m_3\leftarrow d,x_3\in F(m_3)),$

@@

Lastly, we require two globular 2-morphisms $\chi \colon (M_1 \otimes N_1) \odot (M_2 \otimes N_2) \to (M_1 \odot M_2) \otimes (N_1 \odot N_2)$ and $\mu \colon U_{a \otimes b} \to U_a \otimes U_b$, the first of which relates the double functor $\otimes$ with the functor $\odot$ and the second of which relates the double functor $\otimes$ with the functor $U$. The latter of these is easy to see: given two objects $a$ and $b$ in $F\lCsp_0=\A$, $\mu$ is given by the identity map of cospans in $\A$:

@@

together with the decoration isomorphism $\tau_{a,b} \maps \bot_{a+b} \xrightarrow{\sim} \bot_a + \bot_b$ given by the unique map between the two initial objects $\bot_{a+b}$ and $\bot_a + \bot_b$ in $F(a+b)$. For the former globular 2-morphism $\chi$, we first introduce some notation for objects $M_1,M_2,N_1$ and $N_2$ of $F\lCsp_1$:

@@

@@

The first part formulates the existence of colimits \emph{locally} in each fibre, and if we let $\J$ range over all finite categories it says that the corresponding pseudofunctor $F\maps\A\to\Cat$ lands in the sub-2-category $\Rex$. The second part formulates the existence of colimits \emph{globally} in the total category $\inta F$, and if we let $\J$ range over all finite categories it says that $\X$ is finitely cocomplete and $U$ preserves all finite colimits. As an example, suppose that $\A$ has pushouts and a Grothendieck opfibration $U\colon\inta F\to\A$ has fibrewise pushouts preserved by its reindexing functors. We can construct pushouts in $\inta F$ as follows: 
\begin{equation}\label{eq:globalpushout}
 \begin{tikzcd}[ampersand replacement=\&,column sep=1in,row sep=.5in]
\& (a+_bc,w)\ar[dd,phantom, very near start,"\upback"description] \& \\
(a,x)\ar[ur] \&\& (c,z)\ar[ul] \\
\& (b,y)\ar[ul,"{\begin{cases}f\maps b\to a &\textrm{in }\A \\k\maps F(f)y\to x &\textrm{in }F(a)\end{cases}}"description]\ar[ur,"{\begin{cases}g\maps b\to c &\textrm{in }\A \\\ell\maps F(g)y\to z &\textrm{in }F(c)\end{cases}}"description] \&
 \end{tikzcd}
\end{equation}
which is the pushout $\sigma_a\colon a\to a+_bc\leftarrow c:\sigma_c$ in $\A$ on the first variable, and $w$ is the pushout in the fibre $F(a+_bc)$
\begin{displaymath}
\begin{tikzcd}
 & w \ar[dd,phantom, very near start,"\upback"description] & \\
 F(\sigma_a)x\ar[ur] & & F(\sigma_c)z\ar[ul] \\
 & F(\sigma_a)(Ff(y))\cong F(\sigma_c)(Fg(y))\ar[ul,"F(\sigma_a)k"]\ar[ur,"F(\sigma_c)\ell"'] &
 \end{tikzcd}
\end{displaymath}
In similar ways we can construct the initial object and coproducts in a Grothendieck category, namely
\begin{equation}\label{eq:fibrewisecoprod}
(a,x)+(b,y)=\left(a+ b,F(\iota_a)x+F(\iota_b)y\right), \qquad 
 0_{\inta F}=\left(0_A,0_{F(0_A)}\right)
\end{equation}
where $\iota_a:a\to a+b\leftarrow b:\iota_b$ are the sum inclusions in $\A$.

@@

Building up to an isomorphism of the above described double categories, we now define a double functor
\begin{displaymath}
\lE=(\lE_0,\lE_1)\maps _L\lCsp(\inta F) \longrightarrow F\lCsp
\end{displaymath}
which on vertical categories acts as the identity $\lE_0 = \id_{\A}$. Given a horizontal 1-cell in $_L\lCsp(\inta F)$ as in \cref{eq:scsphor1cell}, $\lE_1$ maps it to the decorated cospan $i\maps a\to m\leftarrow b:o$ with decoration $x\in F(m)$,
whereas a 2-morphism of $L$-structured cospans as above is mapped to the underlying map of cospans in $\A$
\begin{displaymath}
\begin{tikzcd}
a\ar[r,"i"]\ar[d,"f"'] & m\ar[d,"k"] & b\ar[l,"o"']\ar[d,"g"] \\
a'\ar[r,"i'"']& m' & b'\ar[l,"o'"]
\end{tikzcd}
\end{displaymath}
in $\A$ along with the map $\tau\maps Fk(x)\to x'$ in $F(m')$, namely an $F$-decorated cospan 2-morphism. 
With these definitions, $\lE$ is indeed a (pseudo) double functor as in \cref{def:doublefun}: both $\lE_0$ and $\lE_1$ are functors that appropriately commute with sources and targets, and there exist globular 2-isomorphisms $\lE(M) \odot \lE(N) \simrightarrow\lE(M \odot N)$, $U_{\lE m} \simrightarrow\lE(U_m)$
for any pair of composable horizontal 1-cells $M=\cspn{(a,\bot_a)}{(m,x)}{(b,\bot_b)}{}{}$ and $N=\cspn{(b,\bot_b)}{(n,y)}{(c,\bot_c)}{}{}$ and any object $m$ of $_L\lCsp(\inta F)$. Explicitly, $\mathbb{E}(M)\odot \mathbb{E}(N)$ is given as in \cref{eq:dcospanscomposition} via the pushout cospan and decoration
\begin{equation}\label{eq:1}
 \begin{tikzcd}
  & m+_bn & \\
  a\ar[ur] && b,\ar[ul]
 \end{tikzcd}\qquad
  \one\xrightarrow{x\times y}F(m)\times F(n)\xrightarrow{\phi_{m,n}}F(m+n)\xrightarrow{F(\psi)}F(m+_bn)
\end{equation}
whereas first composing $M$ and $N$ in $_L\lCsp(\inta F)$ using pushouts in $\inta F$ as in \cref{eq:globalpushout} produces 
\begin{displaymath}
 \begin{tikzcd}[column sep={.7in,between origins}]
 && (m+_b n,F(\sigma_m)x+_{(\bot_{m+_bn})}F(\sigma_n)y) && \\
 & (m,x) \ar[ur] && (n,y)\ar[ul] \\
 (a,\bot_a)\ar[ur] && (b,\bot_b)\ar[ur]\ar[ul] && (c,\bot_c)\ar[ul]
 \end{tikzcd}
\end{displaymath}
where $\sigma$ are the canonical maps into the pushout. %overusing notation, streamline with earlier
Under $\lE$ this is mapped to the underlying $\A$-cospan $\cspn{a}{m+_bn}{b}{}{}$ with decoration $F(\sigma_m)x+F(\sigma_n)y$, since a pushout over an initial object is really a coproduct. That decoration is isomorphic to that of \cref{eq:1} due to the following diagram 
\begin{equation}\label{eq:isomorphismdiag}
 \begin{tikzcd}
F(m)\times F(n)\ar[rr,"\phi"]\ar[d,"F(\sigma_m)\times F(\sigma_n)"']\ar[dr,phantom,"\cong"] && F(m+n)\ar[d,"F(\psi)"]\ar[d,bend right=50, phantom,"\cong"']\ar[dl,"F(\sigma_m+\sigma_n)"description] \\
F(m+_bn)\times F(m+_bn)\ar[r,"\phi"'] & F\left((m+_bn)+(m+_bn)\right)\ar[r,"F(\nabla)"'] & F(m+_bn)
 \end{tikzcd}
\end{equation}
where the bottom composite is the fibrewise coproduct in $F(m+_bn)$ as in \cref{eq:explicitstructure1}. The left-hand side isomorphism is due to pseudonaturality of $\phi$ whereas the right-hand side isomorphism follows from universal properties and pseudofunctoriality of $F$.
Regarding identities, the identity horizontal 1-cell $U_m$ in $_L\lCsp(\inta F)$ is $\cspn{(m,\bot_m)}{(m,\bot_m)}{(m,\bot_m)}{}{}$ with $1_m$ as the $A$-component and isomorphisms between initial objects in the fibres. Thus $\lE(U_m)$ is the identity cospan in $A$ together with the `initial decoration' $\bot_m\in F(m)$, and so is $U_{\mathbb{E}(m)}$ as in \cref{eq:UFCsp}. %Do we want to check that E is a double functor one last time?

@@

{\chris The rest is not done yet...}

>>>>

\begin{tikzpicture}[scale=1.2,baseline=(current bounding box.center)]

@@

(B) edge node [below] {$j_m$} (F)
(D) edge node [below] {$j_{n}$} (F)
(B) edge[bend left=10] node [above] {$u_m\;$} (G)
(D) edge[bend right=10] node [above] {$\;u_n$} (G)
(F) edge node [fill=white] {$\psi$} (G)

@@

is formed as follows: for three composable horizontal 1-cells $M_1=(a\to m_1\leftarrow b,x_1\in F(m_1))$, $M_2=(b\to m_2\leftarrow c,x_2\in F(m_2))$ and $M_3=(c\to m_3\leftarrow d,x_3\in F(m_3)),$

@@

Lastly, according to \cref{defn:monoidal_double_category} we construct two globular 2-isomorphisms $(M_2 \otimes N_2)\odot (M_1 \otimes N_1) \simrightarrow(M_2\odot M_1) \otimes (N_2\odot N_1)$ and $U_{a + b} \simrightarrow U_a \otimes U_b$ that express compatibility of the tensor with horizontal composition and units. For any appropriate four decorated cospans
\begin{gather*}
M_1=\left(a\xrightarrow{i_1}m_1\xleftarrow{o_1}b,x_1\in F(m_1)\right),\;\; M_2=\left(b\xrightarrow{i_2}m_2\xleftarrow{o_2}c,x_2\in F(m_2)\right) \\
N_1=\left(a'\xrightarrow{i'_1}n_1\xleftarrow{o'_1}b',y_1\in F(n_1)\right),\;\; N_2=\left(b'\xrightarrow{i'_2}n_2\xleftarrow{o'_2}c',y_2\in F(n_2)\right)
\end{gather*}
first tensoring $M_1\ot N_1$, $M_2\ot N_2$ by \cref{eq:tensordecoration} and then horizontally composing them by \cref{eq:dcospanscomposition} gives
\begin{displaymath}
\scalebox{.8}{\begin{tikzcd}[column sep={.6in,between origins},ampersand replacement=\&]
\&\& (m_1+n_1)+_{b{+}b'}(m_2+n_2) \&\& \\
\& m_1+n_1\ar[ur] \&\& m_2+n_2\ar[ul] \& \\
a+a'\ar[ur,"{i_1+i'_1}"] \&\& b+b'\ar[ul,"{o_1+o'_1}"']\ar[ur,"{i_2+i'_2}"] \&\& c+c'\ar[ul,"{o_2+o'_2}"']
\end{tikzcd}}
\scalebox{.8}{\begin{tikzcd}[ampersand replacement=\&,column sep=.5in]
1\ar[ddrr,dashed,"{(x_1\oplus y_1)\odot(x_2\oplus y_2)}"']\ar[r,"x_1{\times}y_1{\times}x_2{\times}y_2"] \& F(m_1){\times}F(n_1){\times}F(m_2){\times}F(n_2)\ar[r,"\phi_{m_1,n_1}\times\phi_{m_2,n_2}"] \& F(m_1{+}n_1){\times}F(m_2{+}n_2) \ar[d,"\phi_{m_1+n_1,m_2+n_2}"] \\
\&\& F(m_1+n_1+m_2+n_2)\ar[d,"F\psi"] \\
\&\& F((m_1+n_1)+_{b+b'}(m_2+n_2))
\end{tikzcd}}
\end{displaymath}
If we first horizontally compose $M_2\odot M_1$, $N_2\odot N_1$ and then tensor the result, we obtain
\begin{displaymath}
  \scalebox{.8}{\begin{tikzcd}[column sep=.3in,ampersand replacement=\&]
 \& (m_1+_b m_2)+(n_1+_{b'}n_2) \& \\
 \hole \\
 a+a'\ar[uur]%,"{u_{m_1}i_1+u_{n_1}i'_1}"] 
 \&\& c+c'\ar[uul]%,"{u_{m_2}o_2+u_{n_2}o'_2}"']  
 \end{tikzcd}}
 \scalebox{.8}{\begin{tikzcd}[ampersand replacement=\&,column sep=.4in]
1\ar[ddrr,dashed,"{(x_1\odot x_2)\oplus(y_1\odot y_2)}"']\ar[r,"x_1{\times}x_2{\times}y_1{\times}y_2"] \& F(m_1){\times}F(m_2){\times}F(n_1){\times}F(n_2)\ar[r,"\phi_{m_1,m_2}\times\phi_{n_1,n_2}"] \& F(m_1{+}m_2){\times}F(n_1{+}n_2) \ar[d, "F\psi\times F\psi"] \\
\&\& F(m_1+_bm_2)\times F(n_1+_{b'}n_2)\ar[d,"\phi_{m_1+_bm_2,n_1+_{b'}n_2}"] \\
\&\& F((m_1+_bm_2)+F(n_1+_{b'}n_2))
\end{tikzcd}}
\end{displaymath}
where the legs of the cospan are the sums of the dashed ones of the appropriate diagrams \cref{eq:dcospanscomposition}.
Thus the (globular) cospan isomorphism is the canonical universal map between the colimit of the same diagram obtained in two different ways $(m_1+n_1)+_{b+b'}(m_2+n_2)\cong(m_1+_b m_2)+(n_1+_{b'}n_2)$, and the decoration isomorphism can be built in a similar way using coherent isomorphisms essentially due to pseudonaturality of $\phi$. {\chris Should we give more details here?}
Similarly, for units $U_{a+b}$ and $U_a\ot U_b$ it is easy to see that as cospans they are both the identity $\id:a+b\to a+b\leftarrow a+b:\id$, and there is a canonical isomorphism between their trivial decorations \cref{eq:UFCsp}, namely $F(!_{a+b})\phi_0\cong\phi_{a,b}(F(!_a)\times F(!_b))(\phi_0\times\phi_0)$ via the structure isomorphisms for $(\phi,\phi_0)$.
, the first of which relates the double functor $\otimes$ with the functor $\odot$ and the second of which relates the double functor $\otimes$ with the functor $U$. 
The latter of these is easy to see: given two objects $a$ and $b$ in $F\lCsp_0=\A$, $\mu$ is given by the identity map of cospans in $\A$:

@@

together with the decoration isomorphism $\tau_{a,b} \maps \bot_{a+b} \xrightarrow{\sim} \bot_a + \bot_b$ given by the unique map between the two initial objects $\bot_{a+b}$ and $\bot_a + \bot_b$ in $F(a+b)$. 
%%%Christina: we do not have any such information at this point! We have not talked about the categories F(a) being cocomplete, and it's not deduced from the assumptions. We will show this is the case much later!
For the former globular 2-morphism $\chi$, we first introduce some notation for objects $M_1,M_2,N_1$ and $N_2$ of $F\lCsp_1$:

@@

{\chris IN progress}

@@

The first part formulates the existence of colimits \emph{locally} in each fibre, and if we let $\J$ range over all finite categories it says that the corresponding pseudofunctor $F\maps\A\to\Cat$ lands in the sub-2-category $\Rex$. The second part formulates the existence of colimits \emph{globally} in the total category $\inta F$, and if we let $\J$ range over all finite categories it says that $\X$ is finitely cocomplete and $U$ preserves all finite colimits. As an example, suppose that $\A$ has pushouts and a Grothendieck opfibration $U\colon\inta F\to\A$ has fibrewise pushouts preserved by its reindexing functors. We can construct pushouts in $\inta F$ as follows: 
\begin{equation}\label{eq:globalpushout}
 \begin{tikzcd}[ampersand replacement=\&,column sep=1in,row sep=.5in]
\& (a+_bc,w)\ar[dd,phantom, very near start,"\upback"description] \& \\
(a,x)\ar[ur] \&\& (c,z)\ar[ul] \\
\& (b,y)\ar[ul,"{\begin{cases}f\maps b\to a &\textrm{in }\A \\k\maps F(f)y\to x &\textrm{in }F(a)\end{cases}}"description]\ar[ur,"{\begin{cases}g\maps b\to c &\textrm{in }\A \\\ell\maps F(g)y\to z &\textrm{in }F(c)\end{cases}}"description] \&
 \end{tikzcd}
\end{equation}
which is the pushout $u_a\colon a\to a+_bc\leftarrow c:u_c$ in $\A$ on the first variable, and $w$ is the pushout in the fibre $F(a+_bc)$
\begin{displaymath}
\begin{tikzcd}
 & w \ar[dd,phantom, very near start,"\upback"description] & \\
 F(u_a)x\ar[ur] & & F(u_c)z\ar[ul] \\
 & F(u_a)(Ff(y))\cong F(u_c)(Fg(y))\ar[ul,"F(u_a)k"]\ar[ur,"F(u_c)\ell"'] &
 \end{tikzcd}
\end{displaymath}
In similar ways we can construct the initial object and coproducts in a Grothendieck category, namely
\begin{equation}\label{eq:fibrewisecoprod}
(a,x)+(b,y)=\left(a+ b,F(\iota_a)x+F(\iota_b)y\right), \qquad 
 0_{\inta F}=\left(0_A,0_{F(0_A)}\right)
\end{equation}
where $\iota_a:a\to a+b\leftarrow b:\iota_b$ are the sum inclusions in $\A$.

@@

Building up to an isomorphism of the above described double categories, we now define a double functor
\begin{displaymath}
\lE=(\lE_0,\lE_1)\maps _L\lCsp(\inta F) \longrightarrow F\lCsp
\end{displaymath}
which on vertical categories acts as the identity $\lE_0 = \id_{\A}$. Given a horizontal 1-cell in $_L\lCsp(\inta F)$ as in \cref{eq:scsphor1cell}, $\lE_1$ maps it to the decorated cospan $i\maps a\to m\leftarrow b:o$ with decoration $x\in F(m)$,
whereas a 2-morphism of $L$-structured cospans as above is mapped to the underlying map of cospans in $\A$
\begin{displaymath}
\begin{tikzcd}
a\ar[r,"i"]\ar[d,"f"'] & m\ar[d,"k"] & b\ar[l,"o"']\ar[d,"g"] \\
a'\ar[r,"i'"']& m' & b'\ar[l,"o'"]
\end{tikzcd}
\end{displaymath}
in $\A$ along with the map $\tau\maps Fk(x)\to x'$ in $F(m')$, namely an $F$-decorated cospan 2-morphism. 
With these definitions, $\lE$ is indeed a (pseudo) double functor as in \cref{def:doublefun}: both $\lE_0$ and $\lE_1$ are functors that appropriately commute with sources and targets, and there exist globular 2-isomorphisms $\lE(M) \odot \lE(N) \simrightarrow\lE(M \odot N)$, $U_{\lE m} \simrightarrow\lE(U_m)$
for any pair of composable horizontal 1-cells $M=\cspn{(a,\bot_a)}{(m,x)}{(b,\bot_b)}{}{}$ and $N=\cspn{(b,\bot_b)}{(n,y)}{(c,\bot_c)}{}{}$ and any object $m$ of $_L\lCsp(\inta F)$. Explicitly, $\mathbb{E}(M)\odot \mathbb{E}(N)$ is given as in \cref{eq:dcospanscomposition} via the pushout cospan and decoration
\begin{equation}\label{eq:1}
 \begin{tikzcd}
  & m+_bn & \\
  a\ar[ur] && b,\ar[ul]
 \end{tikzcd}\qquad
  \one\xrightarrow{x\times y}F(m)\times F(n)\xrightarrow{\phi_{m,n}}F(m+n)\xrightarrow{F(\psi)}F(m+_bn)
\end{equation}
whereas first composing $M$ and $N$ in $_L\lCsp(\inta F)$ using pushouts in $\inta F$ as in \cref{eq:globalpushout} produces 
\begin{displaymath}
 \begin{tikzcd}[column sep={.7in,between origins}]
 && (m+_b n,F(u_m)x+_{(\bot_{m+_bn})}F(u_n)y) && \\
 & (m,x) \ar[ur] && (n,y)\ar[ul] \\
 (a,\bot_a)\ar[ur] && (b,\bot_b)\ar[ur]\ar[ul] && (c,\bot_c)\ar[ul]
 \end{tikzcd}
\end{displaymath}
where $u$ are the canonical maps into the pushout.
Under $\lE$ this is mapped to the underlying $\A$-cospan $\cspn{a}{m+_bn}{b}{}{}$ with decoration $F(u_m)x+F(u_n)y$, since a pushout over an initial object is really a coproduct. That decoration is isomorphic to that of \cref{eq:1} due to the following diagram 
\begin{equation}\label{eq:isomorphismdiag}
 \begin{tikzcd}
F(m)\times F(n)\ar[rr,"\phi"]\ar[d,"F(u_m)\times F(u_n)"']\ar[dr,phantom,"\cong"] && F(m+n)\ar[d,"F(\psi)"]\ar[d,bend right=50, phantom,"\cong"']\ar[dl,"F(u_m+u_n)"description] \\
F(m+_bn)\times F(m+_bn)\ar[r,"\phi"'] & F\left((m+_bn)+(m+_bn)\right)\ar[r,"F(\nabla)"'] & F(m+_bn)
 \end{tikzcd}
\end{equation}
where the bottom composite is the fibrewise coproduct in $F(m+_bn)$ as in \cref{eq:explicitstructure1}. The left-hand side isomorphism is due to pseudonaturality of $\phi$ whereas the right-hand side isomorphism follows from universal properties and pseudofunctoriality of $F$.
Regarding identities, the identity horizontal 1-cell $U_m$ in $_L\lCsp(\inta F)$ is $\cspn{(m,\bot_m)}{(m,\bot_m)}{(m,\bot_m)}{}{}$ with $1_m$ as the $A$-component and isomorphisms between initial objects in the fibres. Thus $\lE(U_m)$ is the identity cospan in $A$ together with the `initial decoration' $\bot_m\in F(m)$, and so is $U_{\mathbb{E}(m)}$ as in \cref{eq:UFCsp}. %Do we want to check that E is a double functor one last time?

@@

%{\chris The rest is not done yet...} %%%Christina I have no clue where this comment came from!
